### 2016-11-11 Paper_v1.tex
MVupdates2

% Load packages
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{mathtools}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{fullpage}
\usepackage{comment}
\usepackage{lineno}
%\usepackage{chicago}
\usepackage{textcomp}
\usepackage{multirow}
\usepackage{amsmath}
%\usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
\DeclareMathOperator*{\argmin}{\arg\!\min}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell Free Protein Synthesis in \emph{Escherichia~coli}}}}
\vspace{0.05in}
{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Che-Hsiao Shih$^{1}$, Joseph Wayman, Kara Calhoun$^{2}$, James Swartz$^{2}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering {$^{1}$School of Chemical Engineering}}
{\par \centering {Purdue University, Lafayette, IN 47901}}
\vspace{0.1in}
{\par \centering {$^{2}$School of Chemical Engineering}}
{\par \centering {Stanford University, Stanford, CA 94305}}
{\par \centering \textbf{Running Title:}~Dynamic modeling of cell free protein synthesis}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Cell free protein expression systems have become widely used in systems and synthetic biology.
In this study, we developed an ensemble of dynamic \textit{E. coli} cell free protein synthesis (CFPS) models.
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
Taken together, we presented the first dynamic model of \textit{E. coli} cell free protein synthesis.
This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

\vspace{0.1in}
{\noindent \textbf{Keywords:}~Biochemical engineering, systems biology, cell free protein synthesis}

@@

In this study, we developed an ensemble of \textit{E. coli} cell free protein synthesis (CFPS) models using the hybrid cell free modeling approach of Wayman et al [REFHERE].
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
Taken together, we presented the first dynamic model of \textit{E. coli} cell free protein synthesis.
We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

\subsection*{Estimation of an ensemble of cell free protein synthesis models.}
We used the hybrid cell free modeling framework of Wayman et al. to simulate the production of a model protein [REFHERE].
The cell-free \textit{E. coli} metabolic model was constructed by removing the growth-associated processes from the model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS}, and by adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle),
as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids. We generated an ensemble of N = 18,000 parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model. We defined the set with the lowest error value as the best-fit parameter set. [STATISTICS ON PARAMETERS].

The ensemble of models captured the time evolution of cell free CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
Glucose was exhausted with 3 hr [FILL ME IN].
The ensemble also captured the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:CarbonBoth}, top) and the totals by base .
The ensemble and the best-fit set also predict some of the amino acid measurements, while failing to predict others (Fig.~\ref{fig:Amino}).
the central carbon metabolism, including glucose uptake, CAT production, and the dynamics of the organic acid intermediates .
Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).
This is likely due to a structural deficiency in the model; in some cases, the consumption of an amino acid through CAT synthesis is not enough to explain the decrease shown in the data, and there are no other reactions that consume it. Thus, a more comprehensive biological description is needed to fully explain amino acid dynamics.

@@

\subsection*{Maximum theoretical CAT yield showed CFPS can be optimized.}
We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
%Arginine and glutamate were excluded due to not being present in the training dataset.
The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
While the model ensemble described the experimental data, it was unclear whether the performance of the CFPS system was optimal.
To address this question, we used ssFBA in combination with the cell-free metabolic network and a detailed promoter model under a T7 polymerase to compute the maximum theoretical carbon yield.
However, we first validated the ssFBA approach by comparing an ensemble of simulated versus measured concentrations of CAT over a one hour period (Fig. ~\ref{fig:CATProdYield}A). 
The ensemble of 100 sets captured the CAT concentration profile which was randomly generated by sampling RNA polymerase levels, ribosome levels and elongation rates in a physiological range.
We then used sequence-specific FBA to calculate a theoretical maximum CAT yield under four different cases: unconstrained, limited oxidative phosphorylation, bounded by transcription/translation rates, and bounded by experimental data (Fig. ~\ref{fig:CATProdYield}B).
The theoretical maximum carbon yield of CAT was 0.349 for an unconstrained case and 0.194 for the transcription and translation constrained case.
Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 25\% of the theoretical maximum and 45\% of a theoretical physiological case.
Whereas, the case constrained by experimental data had a carbon yield of 0.062 $\pm$ 0.02, similar to the experimental yield. 
This allowed us to quantify the amount of carbon being diverted to byproducts, and suggests that there is potential for a doubling of CAT production by reducing this diversion of carbon.
In comparing the flux distributions between the unconstrained and constrained cases (Fig. ~\ref{fig:Network}), both constrained cases heavily utilize the Entner–Doudoroff pathway which may be a first viable knockout to increase CAT yield.

\subsection*{Sensitivity analysis on FBA system}
We conducted global sensitivity on certain fluxes' upper bounds within a constraint-based FBA system, with protein export rate as the objective function.

@@

In this study, we developed an ensemble of \textit{E. coli} cell free protein synthesis (CFPS) models using the hybrid cell free modeling approach of Wayman et al [REFHERE].
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
% Taken together, we presented the first dynamic model of \textit{E. coli} cell free protein synthesis.
% We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
% This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

The cell free model ensemble described the training data with the exception of some of the amino acids.
Specifically, adding more reactions that consume amino acids would improve the model's ability to predict those that show a decrase in the experimental data.
Also, including specific transcription and translation steps for CAT would allow us to more accurately model the complexity and the resource cost of protein synthesis.
Another area for future work is to more thoroughly sample parameter space.
For the metabolites in the dataset, initial conditions were fixed at the initial data values.
All other parameters were varied in a manner so as to best fit the dataset.
However, the resulting ensemble may not represent every biological or practical possibility.
In a different region of parameter space, the system could behave differently, including the flux distribution through the network, the accuracy and spread of ensemble fits, the relative sensitivities, and the yield as a percentage of the theoretical maximum.
Testing the model under a variety of conditions could strengthen or challenge the findings of this study.
Further experimentation could also be used to gain a deeper understanding of model performance under a variety of conditions.
Specifically, CAT production performed in the absence of amino acids could inform the system's ability to manufacture them, while experimentation in the absence of glucose or oxygen could shed light on how important they are to protein synthesis, and under which conditions.
Finally, the approach should be extended to other protein products.
CAT is only a test protein used for model identification; the modeling framework, and to some extent the parameter values, should be protein agnostic.
An important extension of this study would be to apply its insights to other protein applications, where possible.

@@

The translation rate was constrained as:
 \begin{equation}\nonumber
	w_{tl} = K_{P}Ribo\frac{v_{Ribo}}{l_{protein}}[mRNA_{ss}]
\end{equation}
where $K_{P}$ is the polysome amplification constant, $Ribo$ is the ribosome concentration,  $v_{Ribo}$ is the elongation rate (amino acids/hr) of the ribosome, $l_{protein}$ is the number of amino acids in the protein of interest, and $mRNA_{ss}$ is the mRNA concentration at steady state determined by the transcription rate divided by the degradation rate of mRNA.  
An ensemble of flux distributions was calculated for 100 sets by randonmly sampling. 
Glucose, oxygen, and amino acids were modeled as being imported into the system, whereas CAT synthesis and metabolite byproduct formation was modeled as an export from the system.
The rest of the network followed a pseudo steady-state asusmption where all other metabolites were not allowed to accumulate; thus, the network could be solved by linear programming.
The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
The solution flux vector was used to calculate the theoretical carbon yield of CAT for four different cases.
%\begin{equation}\label{eqn:yield}
%	Yield=\frac{w_{CAT}\;C_{CAT}}{\sum_{i=1}^{\mathcal{R}}\max(w_{i},0)\;C_{i}}
%\end{equation}
%where $w_{CAT}$ denotes the CAT export flux and $w_{i}$ denotes the import flux of the $i$th substrate.
For the unconstrained case, all rates were left unbounded. 
An ensemble of flux distributions was calculated by randonmly sampling the maximum specific glucose uptake rate from 30 to 40 mM/hr determined from experimental data.
For the limited oxidative phosphorylation case, an ensemble of flux distributions was generated by randomly sampling the maximum specific oxygen uptake rate to from 0 to 10 mM/hr.
For the TXTL case (constrained by transcription and translation rates), an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature as well as the specific glucose uptake rate from 30 to 40 mM/hr. 
For the case where the flux was constrained by experimental data, the lower and upper bounds where data was avaible were randomly sampled in the physiological range of the data in addition to randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature. 
\clearpage

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.75\textwidth]{./Figures/CAT_prod_yield.pdf}
\caption{Sequence specific flux balance analysis of CAT production and yield. A. 95\% confidence interval of the ensemble (light blue region) for CAT concentration versus time. B. Theoretical maximum carbon yield of CAT calcualted by ssFBA for four different cases: unconstrained (black), limited oxidative phosphorylation activity (gray), constrained by transcription and translation rates (light grey), and constrained by experimental data (white). Error bars represent standard deviation of the ensemble.}
\label{fig:CATProdYield}
\end{figure}

>>>>

% Load packages
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{mathtools}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{fullpage}
\usepackage{comment}
\usepackage{lineno}
%\usepackage{chicago}
\usepackage{textcomp}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{textgreek}
%\usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
\DeclareMathOperator*{\argmin}{\arg\!\min}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell-Free Protein Synthesis in \emph{Escherichia~coli}}}}
\vspace{0.05in}
{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Joseph Wayman, Kara Calhoun$^{1}$, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {Robert Frederick Smith School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering {$^{1}$School of Chemical Engineering}}
{\par \centering {Stanford University, Stanford, CA 94305}}
{\par \centering \textbf{Running Title:}~Dynamic modeling of cell-free protein synthesis}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, Robert Frederick Smith School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Cell-free protein expression systems have become widely used in systems and synthetic biology.
In this study, we developed an ensemble of dynamic \textit{E. coli} cell-free protein synthesis (CFPS) models.
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

\vspace{0.1in}
{\noindent \textbf{Keywords:}~Biochemical engineering, systems biology, cell-free protein synthesis}

@@

In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman et al [REFHERE].
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

\subsection*{Estimation of an ensemble of cell-free protein synthesis models.}
We used the hybrid cell-free modeling framework of Wayman et al. to simulate the production of a model protein [REFHERE].
The cell-free \textit{E. coli} metabolic model was constructed by removing the growth-associated processes from the model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS}, and by adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle),
as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids. We generated an ensemble of N = 18,000 parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model. We defined the set with the lowest error value as the best-fit parameter set. [STATISTICS ON PARAMETERS].

The ensemble of models captured the time evolution of cell-free CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
Glucose was exhausted with 3 hr [FILL ME IN].
The ensemble also captured the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:CarbonBoth}, top) and the totals by base .
The ensemble and the best-fit set also predict some of the amino acid measurements, while failing to predict others (Fig.~\ref{fig:Amino}).
the central carbon metabolism, including glucose uptake, CAT production, and the dynamics of the organic acid intermediates .
Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).
This is likely due to a structural deficiency in the model; in some cases, the consumption of an amino acid through CAT synthesis is not enough to explain the decrease shown in the data, and there are no other reactions that consume it. Thus, a more comprehensive biological description is needed to fully explain amino acid dynamics.

@@

\subsection*{Maximum theoretical CAT yield showed CFPS can be optimized.}
We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
%Arginine and glutamate were excluded due to not being present in the training dataset.
The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
While the model ensemble described the experimental data, it was unclear whether the performance of the CFPS system was optimal.
To address this question, we used ssFBA in combination with the cell-free metabolic network and a detailed promoter model under a T7 polymerase to compute the maximum theoretical carbon yield.
However, we first validated the ssFBA approach by comparing an ensemble of simulated versus measured concentrations of CAT over a one hour period (Fig. ~\ref{fig:CATProdYield}A). 
The ensemble of 100 sets captured the CAT concentration profile which was generated by sampling RNA polymerase levels, ribosome levels and elongation rates in a physiological range.
We then used sequence-specific FBA to calculate a theoretical maximum CAT yield under three different cases: unconstrained, constrained by transcription/translation rates, and constrained by transcription/translation rates and measurements (Fig. ~\ref{fig:CATProdYield}B).
The theoretical maximum carbon yield of CAT was 0.35 $\pm$ 0.006 for an unconstrained case and 0.225 $\pm$ 0.03 for the transcription and translation constrained case.
Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 25\% of the theoretical maximum and 38\% of a theoretical physiological case.
Whereas, the case constrained by experimental data had a carbon yield of 0.062 $\pm$ 0.02, similar to the experimental yield. 
In comparing the flux distributions between the unconstrained and constrained cases (Fig. ~\ref{fig:Network}), the constrained cases heavily utilize the first step in the pentose phosphate pathway to generate NADPH.
The majority of the flux continues through the Entner–Doudoroff pathway whereas in the unconstrained case, the majority of flux travels through glycolysis.
In all cases, the energy source comes from oxidative phosphorylation with some from the citric acid cycle.
Whereas, in just the TX/TL case, there is a high flux through fumerate dehydrogenase from aspartic acid uptake. 
In the unconstrained and most constrained case, we see a mixture of acetate and lactate accumulation.
This shows the system is producing NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to supply enough NADH for oxidative phosphorylation.
As a result, high oxidative phosphorylation activity relative to our cell free system leads to an acetate overflow.
This suggests that there is potential for increasing CAT production by reducing this diversion of carbon.
To simulate potential knockouts, we constrained the specific glucose and amino acid uptake rates to the same values as simulated with no knockouts.
In an ssFBA simulation with constrained TX/TL rates, knocking out the \textit{gnd} reaction decreases flux of acetate production but increases flux through \textit{pntAB} which is responsible for regenerating NADPH.
The simulation showed carbon was diverted towards lactate, however since CAT production is constrained by the translation rate, we expected there to be no increase in CAT production.
The decrease in acetate production is a promising result to increase CAT yield. 
A second simulation with a knockout of \textit{gnd} and phosphate acetyltransferase, showed carbon being diverted towards lactate and succinate, however it required a higher flux through oxidative phosphorylation and the TCA cycle to meet the energetic needs of the system.

@@

In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman et al [REFHERE].
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
balance analysis. The observed CAT yield was 25\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
% Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
% We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
% This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

In comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to the kinetic model and experimental measurements, suggests that there is potential for increasing CAT yield in CFPS as well as CFPS performance. 
The theoretical maximum yield of CAT was 0.35 for an unconstrained case and 0.225 for a transcription/translation constrained case. 
Knockouts of \textit{gnd} and phosphate acetyltransferase, show carbon can be diverted away from acetate and potentially towards CAT or other proteins of interest expressed in CFPS. 
The other limitations to address in CFPS would be to enhance the transcription and translation rates since the protein of interest to be expressed is ultimately bounded by these kinetic rates.    
Li et al. have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA and tRNAs \cite{2014_li_PlosOne}. 
Underwood et al. has also shown the increase in ribosome levels does not significantly increase protein yields and rates, however adding elongation factors increased yields by 23\% at 30 minutes\cite{2005_underwood_biotech}.
In addition to improving CFPS performance, Jewett et al. has showed that oxidative phosphorylation operates in cell-free systems and knocking out oxidative phosphorylation reactions is detrimental to protein yield \cite{2008_jewett_molsysbiol}. 
However, it is inconclusive how much oxidative phosphorylation activity there is compared to \textit{in vivo} systems and both of our models suggest oxidative phosphorylation is vital to CAT production. 
Thus, this is a potential place for improvement to optimize CFPS for better performance and protein yield.

@@

The cell-free model ensemble described the training data with the exception of some of the amino acids.
Specifically, adding more reactions that consume amino acids would improve the model's ability to predict those that show a decrase in the experimental data.
Also, including specific transcription and translation steps for CAT would allow us to more accurately model the complexity and the resource cost of protein synthesis.
Another area for future work is to more thoroughly sample parameter space.
For the metabolites in the dataset, initial conditions were fixed at the initial data values.
All other parameters were varied in a manner so as to best fit the dataset.
However, the resulting ensemble may not represent every biological or practical possibility.
In a different region of parameter space, the system could behave differently, including the flux distribution through the network, the accuracy and spread of ensemble fits, the relative sensitivities, and the yield as a percentage of the theoretical maximum.
Testing the model under a variety of conditions could strengthen or challenge the findings of this study.
Further experimentation could also be used to gain a deeper understanding of model performance under a variety of conditions.
Specifically, CAT production performed in the absence of amino acids could inform the system's ability to manufacture them, while experimentation in the absence of glucose or oxygen could shed light on how important they are to protein synthesis, and under which conditions.
Finally, the approach should be extended to other protein products.
CAT is only a test protein used for model identification; the modeling framework, and to some extent the parameter values, should be protein agnostic.
An important extension of this study would be to apply its insights to other protein applications, where possible.

@@

The translation rate was constrained as:
 \begin{equation}\nonumber
	w_{tl} = K_{P}Ribo\frac{v_{Ribo}}{l_{protein}}[mRNA_{ss}]
\end{equation}
where $K_{P}$ is the polysome amplification constant, $Ribo$ is the ribosome concentration,  $v_{Ribo}$ is the elongation rate (amino acids/hr) of the ribosome, $l_{protein}$ is the number of amino acids in the protein of interest, and $mRNA_{ss}$ is the mRNA concentration at steady state determined by the transcription rate divided by the degradation rate of mRNA.

%\begin{equation}\label{eqn:yield}
%	Yield=\frac{w_{CAT}\;C_{CAT}}{\sum_{i=1}^{\mathcal{R}}\max(w_{i},0)\;C_{i}}
%\end{equation}
%where $w_{CAT}$ denotes the CAT export flux and $w_{i}$ denotes the import flux of the $i$th substrate.
An ensemble of 100 sets of flux distributions was calculated for three different cases, unconstrained, constrained by transcription and translation (TX/TL) rates, and  constrained by TX/TL rates and experimental data. 
For the unconstrained case, all rates were left unbounded, except for the specific glucose uptake rate. 
An ensemble of flux distributions was calculated by randonmly sampling the maximum specific glucose uptake rate from 30 to 40 mM/hr determined from experimental data.
For the case constrained by TX/TL rates, an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the elongation rate by RNA polymerase between 20 and 30 nts/sec, and the elongation rate by ribosomes between 1.5 and 3 AA/sec \cite{2005_underwood_biotech, 2016_garamella_acssynbio}.
For the case constrained by TX/TL rates and experimental data, an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature. 
The lower and upper bound constraints were randomly sampled in the physiological range of the experimental noise where data was available, except for CAT flux, which was determined from the transcription and translation rates. 
\clearpage

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.75\textwidth]{./Figures/CAT_prod_yield.pdf}
\caption{Sequence specific flux balance analysis of CAT production and yield. A. 95\% confidence interval of the ensemble (light blue region) for CAT concentration versus time. B. Theoretical maximum carbon yield of CAT calcualted by ssFBA for three different cases: unconstrained except for glucose uptake (black), constrained by transcription and translation rates (grey), and constrained by transcription, translation rates and experimental measurements where available (light grey). Error bars represent standard deviation of the ensemble.}
\label{fig:CATProdYield}
\end{figure}
